Live Functional Programming with Typed Holes

### 2018-06-07 rebuttal.txt

Thank you for your thoughtful reviews. We believe...

# Review A

> The HazelnutLive calculus in the paper is really basic, it doesn't even have 
> fixpoint.

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator). See Siek and Taha [2006] for more on embedding the untyped lambda
calculus into a gradually typed lambda calculus like Hazelnut Live.

That said, we agree that an explicit fix operator in the style of Plotkin's PCF
might be more clear and there is nothing that would complicate its inclusion, 
so if requested, we would be happy to include it in the second phase submission, 
either in the main calculus or as another extension.

> I'm still confused about which version of the language is implemented.

The language of the submitted implementation is the language of Section 3 + 
the extensions in the appendix. We plan to archive a frozen version of that 
implementation as a simple reference implementation of the core calculus of 
this paper.

The first two paragraphs of Section 2 were, we believe, direct in stating that 
the figures in Section 2 were post-processed to use certain language features 
that were not implemented at the time of submission. None of these language 
features are particularly complex or unusual (records, simple pattern matching, 
and syntactic sugar for function definitions and lists), and they are 
orthogonal to the live programming features that are the topic of the paper, 
which are all implemented as described in Section 2 (see the implementation 
screenshots in Fig. 17-18 in the appendix).

We plan to submit a more recent snapshot of the Hazel implementation for 
artifact evaluation that either substantially narrows or eliminates this 
``convenience gap''.

Our approach has been to...

# Review B

(nothing to say, maybe just acknowledge the typos? remove this section entirely?)

# Review C

> the calculus lacks recursion and control-flow. Dealing with these efficiently 
> and providing a good user interface for inspecting the resulting reified values 
> is the hard problem in making such an environment.

## Recursion

See the response to Review A above regarding recursion in the core calculus.

We also considered the topic of recursion in Sec. 2.2 (Example 2: Recursive 
Functions). The UI was designed with recursion in mind in that it allows the 
user to explore the recursive structure of hole closures iteratively as 
described in that section, rather than displaying nested hole closures all at
once, as is the case in the notation used for the core calculus.

The UI displays a "breadcrumb" path from the result to reflect the exploration 
that the user has done (the basic UI can be seen at the bottom of Fig. 17). We 
would be happy to include this more prominently in the second phase submission 
(it requires showing a second, more deeply nested hole environment, so we will
need the additional space available in the second phase submission).

## Control Flow

The sum types extension in the appendix also adds
primitive control flow. With the additional page budget, we will include this
extension in the main body of the second phase submission.

## Novelty

> While, the technical development is sound, it is mostly based prior work (Hazelnut type system), or a variation on gradually typed calculi, and not very surprising. In fact, evaluation of expression with holes is akin to evaluation of expressions with free variables extended with type casts from gradual typing.

...

### 2018-06-07 rebuttal.txt

Thank you for your thoughtful comments, all of which will be incorporated in our
revision.

@@

# Review A

> I'm still confused about which version of the language is implemented.
> Are the screenshots in section 2 real, or only the ones in the appendix?

The submitted implementation supports the language of Section 3 plus the
extensions in the appendix.

The screenshots in Section 2 are mockups, depicting language features that were
not implemented at the time of submission. To create the screenshots in Section
2, we first ran simpler versions of the examples (using encodings based on the
core features), and then we manually post-processed the resulting HTML/CSS of
the editor to obtain the mockups. The screenshots in the Appendix show the
implementation at the time of submission. We will make this explanation even
clearer in our revision.

For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
implementation that either substantially narrows or eliminates the
``convenience gap'' between the core language and the features depicted in
Section 2.

@@

> The HazelnutLive calculus in the paper is really basic, it doesn't even have 
> fixpoint.

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator). See Siek and Taha [2006] for more on embedding the untyped lambda
calculus into a gradually typed lambda calculus like Hazelnut Live.

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested in our revision, we
would be happy to include either in the main calculus or as an extension.

### 2018-06-07 rebuttal.txt

We plan to submit a more recent snapshot of the Hazel implementation for 
artifact evaluation that either substantially narrows or eliminates this 
``convenience gap''.

Our approach has been to...

# Review C

> While, the technical development is sound, it is mostly based prior work 
> (Hazelnut type system), or a variation on gradually typed calculi, and 
> not very surprising. In fact, evaluation of expression with holes is akin 
> to evaluation of expressions with free variables extended with type casts 
> from gradual typing.

We certainly build directly upon the Hazelnut paper, as reviewed in Sec. 3.1, 
but that paper did not define a dynamic semantics at all.

Gradually typed lambda calculi consider only type holes, and cast failure 
immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
we consider not just type holes but also expression holes, which operate not 
merely as free variables, as suggested, but as closures. Closure tracking is 
critical to the user experience in Sec. 2 and to the fill-and-resume feature 
described in Sec. 4.

Closure tracking has a conceptual relationship with prior work on CMTT 
(a non-obvious insight, we believe!), but that work also did not define an
operational semantics nor consider the commutativity property in Sec. 4.

So the novelty is in carefully combining all three of these pieces of prior work 
and then defining the necessary technical infrastructure (e.g. indeterminate 
forms) to be able to prove preservation, progress and commutativity theorems. 
We are not aware of any comparable related work.

### 2018-06-07 rebuttal.txt

Thank you for your thoughtful comments, all of which will be incorporated in our
revision.

# Review A

> I'm still confused about which version of the language is implemented.
> Are the screenshots in section 2 real, or only the ones in the appendix?

The submitted implementation supports the language of Section 3 plus the
extensions in the appendix. The screenshots in the Appendix show the 
implementation at the time of submission.

All live programming features described in Section 2 are implemented. However,
the figures in Section 2 do depict certain language conveniences that were not 
implemented at the time of submission (per the intro of Section 2). To create 
the figures, we first ran simpler versions of the examples (using encodings 
based on the core calculus), and then we manually post-processed the resulting 
HTML/CSS of the editor to depict standard conveniences common to languages like 
Elm and ML, e.g. records, basic pattern matching and syntactic sugar for 
function definitions and lists.

For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
implementation that either substantially narrows or eliminates this 
"convenience gap".

@@

> The HazelnutLive calculus in the paper is really basic, it doesn't even have 
> fixpoint.

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator). See Siek and Taha [2006] for more on embedding the untyped lambda
calculus into a gradually typed lambda calculus like Hazelnut Live.

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested in our revision, we
would be happy to include it either in the main calculus or as an extension.

### 2018-06-07 rebuttal.txt

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested in our revision, we
would be happy to include it either in the main calculus or as an extension.

# Review C

> While, the technical development is sound, it is mostly based prior work 
> (Hazelnut type system), or a variation on gradually typed calculi, and 
> not very surprising. In fact, evaluation of expression with holes is akin 
> to evaluation of expressions with free variables extended with type casts 
> from gradual typing.

Our focus is the language from the Hazelnut paper, but that paper did not define
a dynamic semantics at all. The solution we propose in this paper certainly
builds on prior work (specifically, gradually typed calculi and CMTT), but we
believe the differences are noteworthy.

1. Gradually typed lambda calculi consider only type holes, and cast failure
immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
we consider not just type holes but also expression holes, which operate not
merely as free variables, as suggested, but as closures. Closure tracking is
critical to the user experience in Sec. 2 and to the fill-and-resume feature
described in Sec. 4.

2. Closure tracking has a conceptual relationship with prior work on CMTT
(a non-obvious insight, we believe!), but that work also did not define an
operational semantics nor consider the commutativity property in Sec. 4.

Combining these pieces of prior work with the necessary technical infrastructure
(e.g. indeterminate forms) to be able to prove preservation, progress and
commutativity theorems was not (for us, at least) completely straightforward; we
iterated through several designs before settling on the proposal in the paper.
We are not aware of any comparable related work.

@@

> the calculus lacks recursion and control-flow. Dealing with these efficiently 
> and providing a good user interface for inspecting the resulting reified values 
> is the hard problem in making such an environment.

## Semantics for Recursion and Control Flow

Please see the discussion above regarding recursion.

The sum types extension in the appendix adds primitive control flow. With the
additional page budget, we will include this extension in the main body of the
second phase submission.

## UI for Recursion and Control Flow

We describe our initial UI design considerations for recursion in Sec. 2.2. The
UI allows the user to explore the recursive structure of hole closures
iteratively as described in that section, rather than displaying nested hole
closures all at once, as is the case in the notation used for the core calculus.

Regarding control flow, the UI displays a "breadcrumb" path from the result to
reflect the exploration that the user has done (the basic UI can be seen at the
bottom of Fig. 17). We would be happy to include this more prominently in the
second phase submission (it requires showing a second, more deeply nested hole
environment, so we will need the additional space available in the second phase
submission).

## Effects and Mutation

TODO

### 2018-06-07 rebuttal.txt

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested in our revision, we
would be happy to include it either in the main calculus or as an extension.

# Review C

> While, the technical development is sound, it is mostly based prior work 
> (Hazelnut type system), or a variation on gradually typed calculi, and 
> not very surprising. In fact, evaluation of expression with holes is akin 
> to evaluation of expressions with free variables extended with type casts 
> from gradual typing.

We start from the Hazelnut paper, but that paper did not define
a dynamic semantics at all. The solution we propose in this paper certainly
builds on prior work (specifically, gradual typing and CMTT), but we
believe the differences are noteworthy.

Gradually typed lambda calculi consider only type holes, and cast failure
immediately aborts evaluation. We do not abort immediately (Sec. 2.4). Moreover,
we consider not just type holes but also expression holes, which operate not
merely as free variables, as suggested, but as closures. Closure tracking is
critical to the user experience in Sec. 2 and to the fill-and-resume feature
described in Sec. 4.

Closure tracking has a conceptual relationship with prior work on CMTT
(a non-obvious insight, we believe!), but that work also did not define an
operational semantics nor consider the commutativity property in Sec. 4.

Combining these pieces of prior work with the necessary technical infrastructure
(e.g. indeterminate forms) to be able to prove preservation, progress and
commutativity theorems was not completely straightforward. We are not aware of 
any comparable related work.

@@

The sum types extension in the appendix adds primitive control flow. With the
additional page budget, we will include this extension in the main body of the
second phase submission.

## UI

We describe our initial UI design considerations for recursion in Sec. 2.2. The
UI allows the user to explore the recursive structure of hole closures
iteratively as described in that section, rather than displaying nested hole
closures all at once, as is the case in the notation used for the core calculus.

Regarding , the UI displays a "breadcrumb" path from the result to
reflect the exploration that the user has done (the basic UI can be seen at the
bottom of Fig. 17). We would be happy to include this more prominently in the
second phase submission (it requires showing a second, more deeply nested hole
environment, so we will need the additional space available in the second phase
submission).

### 2018-06-07 rebuttal.txt

Thank you for your thoughtful feedback!

# Review A

> it doesn't even have fixpoint

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator) - see footnote 9 and Siek and Taha [2006].

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested, we would be happy 
to include it either in the main calculus or as an extension.

@@

> I'm still confused about which version of the language is implemented.
> Are the screenshots in section 2 real, or only the ones in the appendix?

The screenshots in the Appendix show the submitted implementation, which 
supports all of the live programming features described in Section 2 for 
the language of Section 3 plus the extensions in the appendix.

For the sake of exposition, the figures in Section 2 do depict certain language
features that were not implemented at submission (as acknowledged in the 
intro of Section 2). To create these figures, we first ran simpler variants of 
the examples (using the usual lambda encoding tricks), and then we manually
edited the resulting HTML/CSS to add in conveniences common to languages like 
ML/Elm, namely record labels, basic pattern matching and syntactic sugar for 
function definitions and lists.

@@

For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
implementation that either substantially narrows or eliminates this 
"convenience gap".

# Review C

> While, the technical development is sound, it is mostly based prior work 
> (Hazelnut type system), or a variation on gradually typed calculi, and 
> not very surprising. In fact, evaluation of expression with holes is akin 
> to evaluation of expressions with free variables extended with type casts 
> from gradual typing.

Only the external type system comes from the Hazelnut paper (reviewed in 3.1). 
That paper did not define a dynamic semantics at all.

Unlike gradually typed lambda calculi, we do not abort immediately on cast
failure (Sec. 2.4). Moreover, we consider not just type holes but also 
expression holes, which operate not merely as free variables, as suggested, 
but as a sort of closure. Closure tracking is critical to the user experience 
throughout Sec. 2 and to the fill-and-resume feature described in Sec. 4.

Closure tracking has a conceptual relationship with the prior work on CMTT (a 
non-obvious and satisfying logical connection, we believe!), but that work 
also did not define a dynamic semantics nor consider non-empty holes nor the 
commutativity property in Sec. 4.

We carefully combined mechanisms from these prior systems and define new 
technical infrastructure (e.g. indeterminate forms) to be able to prove 
preservation, progress and commutativity theorems mechanically. We are not 
aware of an operational semantics with comparable features and metatheory.

### 2018-06-07 rebuttal.txt

Thank you for your thoughtful feedback!

# Review A

> it doesn't even have fixpoint

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
untyped lambda calculus.

@@

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested, we would be happy
to include it either in the main calculus or as an extension.

> I'm still confused about which version of the language is implemented.
> Are the screenshots in section 2 real, or only the ones in the appendix?

The screenshots in the Appendix show the submitted implementation, which 
supports all of the live programming features described in Section 2 for 
the language of Section 3 plus the extensions in Appendix B.

For the sake of exposition, the figures in Section 2 do depict certain language
features that were not implemented at submission (as acknowledged in the 
intro to Section 2). To create these figures, we first ran simpler variants of 
the examples (using typical lambda encoding tricks) in our implementation, and 
then we manually edited the resulting HTML/CSS only to add in conveniences 
common to languages like ML/Elm, namely record labels, basic pattern matching 
and syntactic sugar for function definitions and lists.

For artifact evaluation, we plan to submit a more recent snapshot of the Hazel
implementation that either substantially narrows or eliminates this small 
"convenience gap".

@@

# Review C

> While, the technical development is sound, it is mostly based prior work 
> (Hazelnut type system), or a variation on gradually typed calculi, and 
> not very surprising. In fact, evaluation of expression with holes is akin 
> to evaluation of expressions with free variables extended with type casts

@@

Only the external type system comes from the Hazelnut paper (reviewed in 3.1). 
That paper did not define a dynamic semantics at all.

Unlike gradually typed calculi, we do not abort immediately on cast failure 
(Sec. 2.4). Moreover, we consider not only type holes but also expression 
holes, which operate not merely as free variables (which, as suggested, would
be rather uninteresting), but as metavariable closures. Closure tracking is 
critical to the user experience throughout Sec. 2 and to the fill-and-resume 
feature in Sec. 4.

Closure tracking has a relationship with the prior work on CMTT (a non-obvious 
and satisfying logical connection, we believe!), but that work also did not 
define a dynamic semantics, nor did it consider non-empty holes nor the 
commutativity property in Sec. 4.

We carefully combined mechanisms from these prior systems and defined new 
technical infrastructure (e.g. indeterminate forms) to be able to prove clean 
preservation, progress and commutativity theorems mechanically. We are not 
aware of a dynamic semantics with comparable features and metatheory.

> the calculus lacks recursion and control-flow

Please see the response above regarding recursion. The sum types extension 
in Appendix B adds primitive control flow. With the additional page budget, 
we could move this extension into the main body.

> providing a good user interface for inspecting the resulting reified values 
> is the hard problem in making such an environment

We agree that both theoretical and UI considerations are worthy of substantial 
investigation. This paper focused on laying detailed type-theoretic foundations
motivated by one proof-of-concept UI design and several small but non-trivial 
examples.

This UI was designed explicitly with recursion in mind in that it allows 
the user to explore the recursive structure of hole closures iteratively (see 
Sec 2.2). It does not display nested hole closures all at once, as in the 
notation used in the core calculus. We say a little more below the fold.

Branches blocked on a hole can certainly "blow up" in size. A simple, partial 
solution is to visually collapse such sub-expressions by default. Because we 
define a small step semantics, another approach is to pause evaluation when 
this situation occurs and continue only if explicitly requested. Because we 
leave evaluation order undefined, the user could also explicitly shift the 
"evaluation cursor" elsewhere in such a situation.

### 2018-06-08 rebuttal.txt

# Review A

> it doesn't even have fixpoint

Type holes actually allow you to express general recursion (e.g. via the Y 
combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
untyped lambda calculus.

That said, we agree that an explicit fix operator might be more clear; there is
nothing that would complicate its inclusion, so if requested, we would be happy
to make this small addition, either in the main calculus or as an extension.

> Are the screenshots in section 2 real, or only the ones in the appendix?

@@

The screenshots in the Appendix show the submitted implementation, which 
supports all of the live programming features described in Section 2 for 
the language of Section 3 plus the extensions in Appendix B.

For the sake of exposition, the figures in Section 2 do go beyond the language
of Section 3 (as acknowledged in the intro to Section 2). To create these figures, 
we first ran simpler variants of the examples (using typical lambda encoding 
tricks) in our implementation, and then we manually edited the resulting 
HTML/CSS only to add in conveniences common to languages like ML/Elm, namely 
record labels, basic pattern matching and syntactic sugar for function 
definitions and lists.

@@

Unlike gradually typed calculi, we do not abort immediately on cast failure 
(Sec. 2.4). Moreover, we consider not only type holes but also expression 
holes, which operate not merely as free variables (which, as suggested, would
be rather uninteresting), but as metavariable closures. Closure tracking is 
critical to the user experience throughout Sec. 2 and to the fill-and-resume 
feature in Sec. 4.

Closure tracking has a relationship with the prior work on CMTT (a non-obvious 
and satisfying logical connection, we believe!), but that work also did not 
define a dynamic semantics, nor did it consider non-empty holes nor the 
commutativity property in Sec. 4.

> the calculus lacks recursion and control-flow

Please see above regarding recursion. The sum types extension in Appendix B 
adds primitive control flow. With the additional page budget, we could move 
this extension into the main body.

> providing a good user interface for inspecting the resulting reified values 
> is the hard problemin making such an environment

We agree that both theoretical and UI considerations are worthy of substantial 
investigation. This paper focused on the former, laying detailed type-theoretic
foundations motivated by one proof-of-concept UI design and several small but 
non-trivial examples.

This UI was designed explicitly with recursion in mind in that it allows 
the user to explore the recursive structure of hole closures iteratively (Sec 2.2). 
It does not display nested hole closures all at once, as in the 
notation used in the core calculus. We say a little more below the fold.

### 2018-06-08 rebuttal.txt

Thank you for your thoughtful feedback!

# Review A

> it doesn't even have fixpoint

Type holes actually give you general recursion (e.g. via the Y 
combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
untyped lambda calculus.

That said, we agree that an explicit fix operator might be more clear; if 
requested, we would be happy to make this small + simple addition.

@@

> Are the screenshots in section 2 real, or only the ones in the appendix?

The appendix shows the submitted implementation, which 
supports all of the live programming features in Section 2 for 
the language of Section 3 plus Appendix B.

For the sake of exposition, the figures in Section 2 do go beyond the language
of Section 3 (acknowledged in the intro to Section 2). To create these figures, 
we ran simpler variants of the examples (using typical lambda encodings), then 
manually edited the HTML/CSS only to add conveniences from Elm, namely record 
labels, pattern matching and sugar for functions and lists.

(We plan to submit a more recent snapshot of Hazel that substantially narrows 
or eliminates this small "convenience gap" for artifact evaluation.)

@@

# Review C

> mostly based prior work (Hazelnut type system), or a variation on gradually 
> typed calculi, and not very surprising. In fact, evaluation of expression 
> with holes is akin to evaluation of expressions with free variables extended 
> with type casts

@@

We disagree with this characterization.

Only the external type system comes from Hazelnut (Sec 3.1). That paper did 
not define a dynamic semantics at all.

Unlike gradually typed calculi, we do not abort on cast failure. Furthermore, 
we consider not only type holes but also expression holes, which operate not 
merely as free variables (which, as suggested, would be uninteresting), but 
as metavariable closures. Closure tracking is critical to the features in  
Sections 2 and 4.

Closure tracking has a relationship with CMTT (a non-obvious 
and satisfying logical connection, we believe!), but that work also did not 
define a dynamic semantics, nor did it consider non-empty holes or the 
commutativity property.

> calculus lacks recursion and control-flow

Please see above regarding recursion. The sum types extension in Appendix B 
adds control flow. We plan to move this into the main body.

> providing a good user interface for inspecting the resulting reified values 
> is the hard problem

Both theoretical and UI considerations are worthy of substantial 
investigation. This paper focuses on the former, motivated by a  
proof-of-concept UI design and several small but non-trivial examples.

This UI was designed with recursion in mind -- it allows you to  
to explore the recursive structure of closures iteratively (Sec 2.2). 
It does not display nested closures all at once, as in the 
formal notation. [more below the fold]

Branches blocked on a hole can certainly "blow up" in size. A partial 
solution is to visually collapse such subexpressions by default. Because we 
define a small step semantics, another approach is to pause evaluation when 
this occurs, continuing only if requested or, because we leave evaluation order
undefined, allowing the user to shift the "evaluation cursor" elsewhere.

>> effects and mutation

The interactions between gradual typing and mutation are well-studied and 
the standard approach (treating refs as invariant wrt consistency) should 
be applicable to recover the properties of Section 3. However, the 
commutativity theorem in Section 4 would not hold.

We anticipate that future work will explore variants of fill-and-resume
that support a weaker commutativity property (e.g. via checkpointing).

However, in order to provide the simplest and strongest theoretical
guarantees, we did not include mutation. (Note that Hazel is 
based on Elm, which is pure and quite practical.)
Thank you for your thoughtful feedback!

Type holes actually give you general recursion (e.g. via the Y 
combinator) - see footnote 9 and [Siek and Taha, 2006] on embedding the 
untyped lambda calculus.

That said, we agree that an explicit fix operator might be more clear; if 
requested, we would be happy to make this small + simple addition.

The appendix shows the submitted implementation, which 
supports all of the live programming features in Section 2 for 
the language of Section 3 plus Appendix B.

For the sake of exposition, the figures in Section 2 do go beyond the language
of Section 3 (acknowledged in the intro to Section 2). To create these figures, 
we ran simpler variants of the examples (using typical lambda encodings), then 
manually edited the HTML/CSS only to add conveniences from Elm, namely record 
labels, pattern matching and sugar for functions and lists.

(We plan to submit a more recent snapshot of Hazel that substantially narrows 
or eliminates this small "convenience gap" for artifact evaluation.)

> mostly based prior work (Hazelnut type system), or a variation on gradually 
> typed calculi, and not very surprising. In fact, evaluation of expression 
> with holes is akin to evaluation of expressions with free variables extended 
> with type casts

Only the external type system comes from Hazelnut (Sec 3.1). That paper did 
not define a dynamic semantics at all.

Unlike gradually typed calculi, we do not abort on cast failure. Furthermore, 
we consider not only type holes but also expression holes, which operate not 
merely as free variables (which, as suggested, would be uninteresting), but 
as metavariable closures. Closure tracking is critical to the features in  
Sections 2 and 4.

Closure tracking has a relationship with CMTT (a non-obvious 
and satisfying logical connection, we believe!), but that work also did not 
define a dynamic semantics, nor did it consider non-empty holes or the 
commutativity property.

> calculus lacks recursion and control-flow

Please see above regarding recursion. The sum types extension in Appendix B 
adds control flow. We plan to move this into the main body.

> providing a good user interface for inspecting the resulting reified values 
> is the hard problem

Both theoretical and UI considerations are worthy of substantial 
investigation. This paper focuses on the former, motivated by a  
proof-of-concept UI design and several small but non-trivial examples.

This UI was designed with recursion in mind -- it allows you to  
to explore the recursive structure of closures iteratively (Sec 2.2). 
It does not display nested closures all at once, as in the 
formal notation. [more below the fold]

Branches blocked on a hole can certainly "blow up" in size. A partial 
solution is to visually collapse such subexpressions by default. Because we 
define a small step semantics, another approach is to pause evaluation when 
this occurs, continuing only if requested or, because we leave evaluation order
undefined, allowing the user to shift the "evaluation cursor" elsewhere.

>> effects and mutation

The interactions between gradual typing and mutation are well-studied and 
the standard approach (treating refs as invariant wrt consistency) should 
be applicable to recover the properties of Section 3. However, the 
commutativity theorem in Section 4 would not hold.

We anticipate that future work will explore variants of fill-and-resume
that support a weaker commutativity property (e.g. via checkpointing).

However, in order to provide the simplest and strongest theoretical
guarantees, we did not include mutation. (Note that Hazel is 
based on Elm, which is pure and quite practical.)

### 2018-06-30 calculus.tex
fixed typos identified by OOPSLA reviews

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation in terms of type assignment as follows:
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

Dynamically, the correctness of fill-and-resume depends on
the following \emph{commutativity} property: if there is some sequence of
steps that go from $d_1$ to $d_2$, then one can fill a hole in these
terms at \emph{either} the beginning or at the end of that step
sequence.
%
We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
transitive closure of stepping (see \Figref{fig:multi-step} in
Appendix~\ref{sec:additional-defns}).
%
\begin{thm}[Commutativity]
  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and $\multiStepsTo{\dexp_1}{\dexp_2}$
  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
%
Critically, this property relies on the more version of
the semantics from \Secref{sec:calculus} where evaluation order is
unspecified (i.e. formally, we omit the bracketed premises).
%
In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
reduce sub-expressions in the same order as a ``fresh'' left-to-right
reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
%
In other words, this notion of fill-and-resume only works for
languages where evaluation order ``does not matter''.
(There are various standard ways to formalize this intuition.
For the sake of space, we review these in Appendix~\ref{sec:confluence}.)

>>>>

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures,~\rulename{TAEHole} and \rulename{TANEHole}, each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation in terms of type assignment as follows:
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

Dynamically, the correctness of fill-and-resume depends on
the following \emph{commutativity} property: if there is some sequence of
steps that go from $d_1$ to $d_2$, then one can fill a hole in these
terms at \emph{either} the beginning or at the end of that step
sequence.
%
We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
transitive closure of stepping (see \Figref{fig:multi-step} in
Appendix~\ref{sec:additional-defns}).
%
\begin{thm}[Commutativity]
  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and $\multiStepsTo{\dexp_1}{\dexp_2}$
  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
%
Critically, this property relies on the version of
the semantics from \Secref{sec:calculus} where evaluation order is
unspecified (i.e. where we omit the bracketed premises).
%
In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
reduce sub-expressions in the same order as a ``fresh'' left-to-right
reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
%
In other words, this notion of fill-and-resume only works for
languages where evaluation order ``does not matter''.
(There are various standard ways to formalize this intuition.
For the sake of space, we review these in Appendix~\ref{sec:confluence}.)
